Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-14 ms.tex
Incorporated Martin's edits

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung radiation in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated to these
temperatures primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two competing theories seeking to explain the radio halo
emission feature the ``hadronic model'' where the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas protons from the ICM \citep[][and references
therein]{article:EnsslinPfrommerMiniatiSubramanian:2011} and the ``re-acceleration model'' where a
CR electron population within the ICM at energies around 100 MeV might be sufficiently long lived so
that during states of powerful ICM turbulence, e.g., after a cluster merger, non-resonant
interactions of these CR electrons with plasma waves might be efficient enough to accelerate them to
high enough energies ($\sim 10$~GeV) for them to emit observable radio emission
\citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of possible
non-thermal emission from clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998}
and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004,
article:Eckert_etal:2007} may provide further indications of relativistic particle populations in clusters,
although the interpretation of these observations as non-thermal diffuse emission have been disputed on
the basis of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR hadrons,
in the form of protons and heavy ions, in the ICM is to be expected. The main energy-loss
mechanism for CR hadrons at high energies is pion production through the interaction of CRs with
nuclei in the ICM. Pions are short lived and decay; the charged pions to electrons and
positrons,\footnote{A charged pion decays to a muon or antimuon, which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to gamma rays. Due to the low density of the
ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large scales of a cluster and the volume
filling magnetic fields in the ICM, CR hadrons will be confined in the cluster on timescales
comparable and longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the
thermal electron pressure. This could then severely jeopardize the use of clusters to determine
cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy
clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs,
magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ --
the radius within which the mean density is 500 times the critical density -- suggesting again a substantial
non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy
gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that
is most sensitive to the cluster core region. However, it assumes that the CR component is fully
mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal
ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux
observable with current generation gamma-ray telescopes rendering this technique at least as
sensitive as the dynamical and hydrostatic method (which are however more general in that they are
sensitive to any non-thermal pressure component).

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example
CMB photons, on ultra-relativistic leptons. Those leptons can either be secondaries from the above
mentioned CR interactions or be injected into the ICM by powerful cluster members and be further
accelerated by diffusive shock acceleration or turbulent reacceleration processes.

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM
particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of
the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004,
article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is
expected to be proportional to the square of the DM density integrated along the line of sight,
making clusters of galaxies potentially good candidates for DM searches
\citep{article:PinzkePfrommerBergstrom2009, article:Cuesta_etal:2011}. This despite their large
distances compared to other common targets for DM searches, such as dwarf spheroidal galaxies
\citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009} or the
Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006,
article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters
\citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a
giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral
radio relic, which appears connected to the radio halo with a ``diffuse'' bridge 
\citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an
infall shock \citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological
simulations that model the non-thermal emission processes
\citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009,
article:Skillman_etal:2011}. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT
all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations
with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo with
substantial turbulent pressure support that shows a lower limit of $\sim 10 \%$ of the total
intracluster medium pressure \citep{article:Schuecker_etal:2004}. As such it is a natural candidate
for gamma-ray observations.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
IACTs \citep{article:Holder_etal:2006} located at an altitude of 1268 m a.s.l. at the Fred Lawrence Whipple
Observatory in southern Arizona (31$^{\circ}$~40\arcmin~30\arcsec~N,
110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes is equipped with a 499 pixel camera
covering a 3.5$^{\circ}$ field of view. The array, completed in the spring of 2007, is designed to detect
gamma-ray emission from astrophysical objects in the energy range from 100 GeV to more than 30 TeV.
The energy resolution is about 15\% and the angular resolution (68\% containment) is about 0.1$^{\circ}$
per event at 1 TeV. The sensitivity of the array allows for detection of a point source with a flux of 1\% of
the steady Crab Nebula flux above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The
integral flux sensitivity above 300 GeV was improved by about 30\% with the move of one telescope in
the summer of 2009. The move had no impact on this data set.}

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray
emission: the core region, the peripheral radio relic, as well as individual powerful cluster
member galaxies. VERITAS has a large enough field of view to allow investigation of several of
these scenarios. In this work the focus has been on the core region and three cluster members. The
latter are treated as point sources in the analysis, whereas the core is treated as both a point
source and a mildly extended source. The intrinsic source extension is assumed to be $0.2^{\circ}$
and $0.4^{\circ}$, based on the extension of the thermal soft X-ray emission from the core. There
is evidence of a recent merger event between the two central galaxies NGC 4889 and NGC 4874
\citep{article:Tribble:1993}. There is also evidence for an excess of non-thermal X-ray emission
from these galaxies as well as the galaxy NGC 4921 \citep{article:Neumann_etal:2003}. Therefore,
searches for point-like VHE gamma-ray emission have been conducted at the locations of these
galaxies. The regions of interest considered in this work are summarized in Table \ref{table:roi}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution in
the Coma cluster with the VERITAS and \emph{Fermi}-LAT upper limits from this work. This approach
includes (1) a simplified multi-frequency analytical approach that assumes a constant CR-to-thermal
energy density and a power-law spectrum in momentum, (2) an analytic model derived from cosmological
hydrodynamical simulations of the formation of galaxy clusters, and (3) an approach that uses the
observed luminosity of the giant radio halo in Coma to place a lower limit on the expected
$\gamma$-ray flux in the hadronic model -- where the radio-emitting electrons are secondaries from
CR interactions and which is independent of the magnetic field distribution.  This translates into a
minimum CR pressure which, if challenged by tight gamma-ray limits/detections, enables scrutinizing
the hadronic interaction model of the formation of giant radio halos. Alternatively, by realizing a
spatial CR distribution that is just consitent with the flux upper limits and requiring the model to
match the observed radio data, this enables us to derive a lower limit on the magnetic field
distribution. This approach, however, assumes the validity of the hadronic interaction model in
explaining the radio halo data.  Modelling the CR distribution through different techniques enable
us to bracket our ignorance about the underlying plasma physics that shape the CR distribution and
hence reflect the Bayesian priors that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it
within plausible ranges of $2.1<\alpha_{\rmn{p}}<2.5$. This spectral range is compatible with the
radio spectral index of the giant radio halo of the Coma cluster after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}.\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.} To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints_simple} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
limits within 0.2$^{\circ}$ are the most constraining, yielding limits on $X_\CR$ between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

\subsection{Simulation-based approach}

We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions that dominates over the inverse Compton (IC) emission from primary and
secondary electrons above 100 MeV in clusters.  Given a density profile as, e.g., inferred by
cosmological simulations or X-ray observations, the analytic approach models the CR distribution and
the associated radiative emission processes from radio to the gamma-ray band. This formalism was
derived from high-resolution simulations of clusters of galaxies that included radiative hydrodynamics,
star formation, supernova feedback, and followed CR physics by tracing the most important injection
and loss processes self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin:2000, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

The overall normalization of the CR and gamma-ray distribution scales (non-linearly) with the
maximum acceleration efficiency at structure formation shock waves. Following recent observations at
supernova remnants \citep{article:Helder_etal:2009} as well as theoretical studies
\citep{article:KangJones:2005}, we adopt an optimistic but nevertheless realistic value of this
parameter and assume that 50\% of the dissipated energy at strong shocks is injected into CRs while
this efficiency rapidly decreases for weaker shocks. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003}, they do not significantly contribute to the CR population in
clusters. Instead, strong shocks during the formation epoch of clusters and strong accretion shocks
at the present time (at the boundary of voids and filaments/super-cluster regions) dominate the
injection of CRs which are adiabatically transported through the cluster. Hence, these model
predictions provide a plausible upper limit for the CR contribution from structure formation shocks
in galaxy clusters which can be scaled with the effective acceleration efficiency. Other possible CR
sources, such as AGN and starburst-driven galactic winds have been neglected for simplicity which
could in principle increase the expected gamma-ray yield.

These cosmological simulations only considered advective transport of CRs by turbulent gas motions
which produces centrally enhanced profiles.  However, other means of CR transport such as diffusion
and streaming may flatten the CR radial profiles. It has been suggested that advection is dominant
only during mergers, in which case relaxed clusters could have a reduced gamma-ray luminosity by up
to a factor of five compared with merging clusters
\citep{article:EnsslinPfrommerMiniatiSubramanian:2011}. The goal of this and other future work of
seeking for gamma-ray emission from clusters is to constrain the acceleration physics and transport
properties of CRs.

We adopt the electron density profile as in \S \ref{sec:simple} and model the temperature profile
of the Coma cluster with a constant central temperature of $kT= 8.25$~keV and a characteristic
decline toward the cluster periphery in accordance with a fit to the universal profile derived from
cosmological cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007} that
also follows the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{article:Vikhlinin_etal:2005}. This enables us to adopt the spatial and spectral distribution
of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution
of cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
model fluxes are compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistic value of the shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

>>>>

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated
primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two theories competing to explain the radio halo
are the ``hadronic model'', in which the radio-emitting electrons are produced in inelastic
collisions of cosmic-ray (CR)  ions with thermal gas of the ICM \citep{article:Dennison:1980},
%\citep[][and references therein]{article:EnsslinPfrommerMiniatiSubramanian:2011}
and the ``re-acceleration model'', in which during states of strong ICM turbulence, e.g. after a
cluster merger, interactions of a long-lived pool of 100-MeV electrons with plasma waves might
be efficient enough to accelerate them to energies ($\sim 10$ GeV) sufficient to produce observable
radio emisson \citep[][and references therein]{article:BrunettiLazarian:2010}. Recent observations of
possibly non-thermal emission from clusters in the the extreme ultraviolet
\citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays \citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may provide further indications of
relativistic particle populations in clusters, although the interpretation of these observations as non-thermal diffuse emission have been disputed on the basis of more sensitive observations
\citep[see, e.g.,][]{article:Ajello_etal:2009, article:Ajello_etal:2010, article:Wik_etal:2009}.

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration
in the ICM is an efficient process, a population of highly relativistic CR protons and heavy ions is to
be expected in the ICM. The main energy-loss mechanism for CR hadrons at high energies is pion
production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay; the
charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon,
which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to
gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large
scales of a cluster and the volume filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable to or longer than the Hubble time \citep[][]{article:Volk_etal:1996,
article:Berezinsky_etal:1997} allowing them to accumulate over time. Depending on the CR pressure,
quantified by the CR-to-thermal pressure fraction
$X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays
can be expected.

A substantial value of $X_\CR$ can bias hydrostatic estimates of cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect. This could then severely jeopardize the use of clusters to
determine cosmological parameters. Comparing X-ray and optical potential profiles in the centers of
galaxy clusters yields an upper limit of 20-30 \% of non-thermal pressure (that can be composed of
CRs, magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic mass estimate compared to the lensing mass of $20 \%$ at
$R_{500}$ -- the radius within which the mean density is 500 times the critical density --
suggesting again a substantial non-thermal pressure contribution on large scales
\citep{article:Mahdavi_etal:2008}. Observing high-energy gamma-ray emission is a complementary
method of constraining the pressure contribution of CRs that is most sensitive to the cluster core
region. However, it assumes that the CR component is fully mixed with the ICM and may not a
two-phase structure of CRs and the thermal ICM. A fraction $X_\CR$ of only a few percent is
required in order to produce a gamma-ray flux observable with current generation gamma-ray
telescopes rendering this technique at least as sensitive as the dynamical and hydrostatic method
(which are more general in that they are sensitive to any non-thermal pressure component).

Gamma-ray emission can also be produced by Compton up-scattering of ambient photons, for example
CMB photons, on ultra-relativistic leptons. Those leptons can either be secondaries from the above
mentioned CR interactions or be injected into the ICM by powerful cluster members and be further
accelerated by diffusive shock acceleration or turbulent reacceleration processes
\citep[][and references therein]{article:SchlickeiserSieversThiemann:1987}.

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM
particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of
the cluster mass is in the form of dark matter. The gamma-ray emissivity from DM annihilations is
proportional to the square of the DM density integrated along the line of sight, making galaxy clusters
potentially good candidates for DM searches \citep{article:EvansFerrerSarkar:2004,
article:BergstromHooper:2006, article:PinzkePfrommerBergstrom2009, article:Cuesta_etal:2011},
despite their large distances compared to other common targets for DM searches, such as dwarf
spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009}
or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006,
article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all
wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc
\citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters
\citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a
giant radio halo \citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral
radio relic, which appears connected to the radio halo with a ``diffuse'' bridge 
\citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested
\citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological simulations, which
model the non-thermal emission processes, \citep{article:PfrommerEnsslinSpringel:2008,
article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}, that the relic is
an infall shock. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky
survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992}. Observations with
XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo supported by
substantial turbulent pressure of at least $\sim 10 \%$ of the total pressure
\citep{article:Schuecker_etal:2004}. The Coma cluster is a natural candidate for gamma-ray
observations.

@@

\section{VERITAS Observations and Analysis}
The VERITAS gamma-ray detector \citep{article:Weekes_etal:2002} is an array of four 12 m diameter
imaging atmospheric Cherenkov telescopes \citep[IACTs;][]{article:Holder_etal:2006} located at an
altitude of 1268 m a.s.l. at the Fred Lawrence Whipple Observatory in southern Arizona
(31$^{\circ}$~40\arcmin~30\arcsec~N, 110$^{\circ}$~57\arcmin~07\arcsec~W). Each of the telescopes
is equipped with a 499 pixel camera covering a 3.5$^{\circ}$ field of view. The array, completed in
the spring of 2007, is designed to detect gamma-ray emission from astrophysical objects in the
energy range from 100 GeV to more than 30 TeV. The energy resolution is about 15\% and the angular
resolution (68\% containment) is about 0.1$^{\circ}$ per event at 1 TeV. The sensitivity of the
array allows for detection of a point source with a flux of 1\% of the steady Crab Nebula flux
above 300 GeV at the $5\sigma$ level in under 30 hours.\footnote{The integral flux sensitivity
above 300 GeV was improved by about 30\% with the move of one telescope in the summer of 2009. The
move had no impact on this data set.}

@@

The Coma cluster is a very rich cluster of galaxies with many plausible sites for gamma-ray
emission: the core region, the peripheral radio relic, as well as individual powerful cluster
member galaxies. VERITAS has a large enough field of view to allow investigation of several of
these scenarios. In this work the focus has been on the core region and three cluster members. The
latter are treated as point sources in the analysis, whereas the core is treated as either a point
source or a mildly extended source with intrinsic source radius $0.2^{\circ}$ or $0.4^{\circ}$,
similar to the extension of the thermal soft X-ray emission from the core. There
is evidence of a recent merger event between the two central galaxies NGC 4889 and NGC 4874
\citep{article:Tribble:1993}. There is also evidence for an excess of non-thermal X-ray emission
from these galaxies as well as the galaxy NGC 4921 \citep{article:Neumann_etal:2003}. Therefore,
searches for point-like VHE gamma-ray emission have been conducted at the locations of these
galaxies. The regions of interest considered in this work are summarized in Table \ref{table:roi}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution
in the Coma cluster using the upper limits dereived from VERITAS and \emph{Fermi}-LAT data in this
work. This approach includes (1) a simplified multi-frequency analytical approach that assumes a
constant CR-to-thermal energy density and a power-law spectrum in momentum, (2) an analytic model
derived from cosmological hydrodynamical simulations of the formation of galaxy clusters, and (3)
an approach that uses the
observed luminosity of the giant radio halo in Coma to place a lower limit on the expected
$\gamma$-ray flux in the hadronic model -- where the radio-emitting electrons are secondaries from
CR interactions and which is independent of the magnetic field distribution.  This translates into a
minimum CR pressure which, if challenged by tight gamma-ray limits/detections, enables scrutinizing
the hadronic interaction model of the formation of giant radio halos. Alternatively, by realizing a
spatial CR distribution that is just consitent with the flux upper limits and requiring the model to
match the observed radio data, this enables us to derive a lower limit on the magnetic field
distribution. We stress again that his approach assumes the validity of the hadronic interaction
model. Modelling the CR distribution through different techniques enable
us to bracket our ignorance about the underlying plasma physics that shape the CR distribution and
hence reflect the Bayesian priors that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

@@

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it in
the range $2.1<\alpha_{\rmn{p}}<2.5$, which is compatible with the
radio spectral index of the giant radio halo of the Coma cluster after accounting for the spectral
steepening at frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}.\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.} To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints_simple} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
flux measurements within 0.2$^{\circ}$ are the most constraining and yield limits on $X_\CR$
between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

@@

\subsection{Simulation-based approach}
We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions which in clusters dominates over the inverse Compton (IC) emission above
100 MeV. Given a density profile as, e.g., inferred by
cosmological simulations or X-ray observations, the analytic approach models the CR distribution and
the associated radiative emission processes from radio to the gamma-ray band. This formalism was
derived from high-resolution simulations of clusters of galaxies that included radiative hydrodynamics,
star formation, supernova feedback, and followed CR physics by tracing the most important injection
and loss processes self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin:2000, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

The overall normalization of the CR and gamma-ray distribution scales non-linearly with the
acceleration efficiency at structure formation shock. Following recent observations at
supernova remnants \citep{article:Helder_etal:2009} as well as theoretical studies
\citep{article:KangJones:2005}, we adopt an optimistic but nevertheless realistic value of this
parameter and assume that 50\% of the dissipated energy at strong shocks is injected into CRs while
this efficiency rapidly decreases for weaker shocks. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003}, they do not significantly contribute to the CR population in
clusters. Instead, strong shocks during the formation epoch of clusters and strong accretion shocks
at the present time (at the boundary of voids and filaments/super-cluster regions) dominate the
injection of CRs which subsequently are adiabatically transported through the cluster. Hence, the model
provides a plausible upper limit for the CR contribution from structure formation shocks
in galaxy clusters which can be scaled with the effective acceleration efficiency. Other possible CR
sources, such as AGN and starburst-driven galactic winds have been neglected for simplicity which
could in principle increase the expected gamma-ray yield.

These cosmological simulations only considered advective transport of CRs by turbulent gas motions
which lead to centrally enhanced density profiles.  However, other means of CR transport such as diffusion
and streaming may flatten the CR radial profiles. It has been suggested that advection is dominant
only during mergers, in which case relaxed clusters could have a reduced gamma-ray luminosity by up
to a factor of five compared with merging clusters
\citep{article:EnsslinPfrommerMiniatiSubramanian:2011}. The goal of this and other future work of
seeking for gamma-ray emission from clusters is to constrain the acceleration physics and transport
properties of CRs.

We adopt the density profile of thermal electrons as discussed in \S \ref{sec:simple} and model the
temperature profile
of the Coma cluster with a constant central temperature of $kT= 8.25$~keV and a characteristic
decline toward the cluster periphery in accordance with a fit to the universal profile derived from
cosmological cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007}
and the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{article:Vikhlinin_etal:2005}. This enables us to adopt the spatial and spectral distribution
of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution
of cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
model fluxes are compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistically large shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the relative CR pressure, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$ as a function of radial distance, $R$, from the Coma cluster
center (red lines) and contained within $R$ (blue lines). All radii are shown in units of the virial
radius, $R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_p c$ to reflect tge high cooling rates at low CR energies. The relative CR
pressure rises toward the outer regions on account of the higher efficiency of CR acceleration at the
peripheral accretion shocks compared to the weak central flow shocks. Adiabatic compression of a
mixture of CRs and thermal gas disfavours the
CR pressure relative to the thermal pressure on account of the softer equation of state of CRs. The
weak increase of $X_{\CR}$ toward the core is due to the comparably fast thermal cooling of gas.
